---
title: 'Statistical Analysis on Diet Plans'
author: 'Anirudh Dahiya'
output:
  html_document:
    toc: yes
    toc_depth: 3
editor_options: 
  chunk_output_type: console
---

---

```{r setup, message=FALSE}
library(tidyverse)
library(dplyr)
options(width=100)
```

# Section 1

#Data Ingestion
```{r Reading Dataset, message = FALSE}
#Reading dataset
diet_data <- read_csv("adult_diet_data.csv", guess_max = 10000)

#Reviewing dataset
str(diet_data)

#Converting 'sex' and 'diet_cond' field values into factors since they're categorical variables
diet_data$sex <- as.factor(diet_data$sex)
diet_data$diet_cond <- as.factor(diet_data$diet_cond)
str(diet_data)
summary(diet_data)

#Removing Outlier - Data cleansing
#Upon applying the summary function on the dataset an outlier can clearly be noticed since the maximum value for the height column has been identified as - 1636.3cm which is not realistic/possible. 
diet_data_processed <- filter(diet_data, diet_data$height != '1636.3')
#Other columns in the dataset seem to contain appropriate data, hence not plotting the dataset as a histogram to check for more outliers

#For renaming axes in plots. 
Labels <- c('M' = "Males",'F' = "Females", 'no_chocolate'="No Chocolate Diet", 'regular'="Regular Diet")

```


----

# Data Dictionary

The following table describes the columns present in the dataset and the corresponding description elaborating on the kind of data present in each column. 

Variable       | Description
-------------- | ---------------------------------------------------------
height         | Height of sample member in cm
start_weight   | Weight in kg of sample member before trial
end_weight     | Weight in kg of sample member at the end of trial
age            | Age of sample member
sex            | Gender of sample member
impulsivity    | Score of sample member on an impulsivity test 
diet_cond      | Type of diet plan
iq             | IQ of sample member

----

#Question 1
```{r Impulsivity & Age Relationship, message = FALSE}
#R code associated for the first request in the assignment
#Plotting a scatterplot for the overall dataset
( ques1_plot1 <- ggplot(data = diet_data_processed, mapping = aes(x = age, y = impulsivity, color = sex)) + 
  geom_point() + 
  geom_smooth(se=FALSE)+
  labs(x="Age", y="Impulsivity", title = "Relatonship Between Age & Impulsivity For Entire Data")) 

#Plotting scatterplots for each gender 
( ques1_plot2 <- ggplot(data = diet_data_processed, mapping = aes(x = age, y = impulsivity, color = sex)) + 
  geom_point() + 
  geom_smooth(se=FALSE)+
  facet_wrap(~sex, nrow =2, labeller = as_labeller(Labels))+
  labs(x="Age", y="Impulsivity", title = "Relatonship Between Age & Impulsivity By Gender") )
```

---

#Question 2
```{r Summary Statistics for BMI, message = FALSE}
#R code associated for the second request in the assignment
#Adding required calculated columns - 'starting_bmi', 'ending_bmi', 'total_weight_loss' to processed dataset
( diet_data_processed <- mutate(diet_data_processed, starting_bmi = diet_data_processed$start_weight/(diet_data_processed$height*0.01)^2, ending_bmi = diet_data_processed$end_weight/(diet_data_processed$height*0.01)^2, total_weight_loss = diet_data_processed$start_weight - diet_data_processed$end_weight) ) 

#Summary statistics for starting BMI
( start_bmi_calc_values <- diet_data_processed %>% group_by(sex, diet_cond) %>% summarise(mean = mean(starting_bmi), sd = sd(starting_bmi)) )

#Summary statistics for ending BMI
( end_bmi_calc_values <- diet_data_processed %>% group_by(sex, diet_cond) %>% summarise(mean = mean(ending_bmi), sd = sd(ending_bmi)) )

#Summary statistics for weight loss
( weight_loss_calc_values <- diet_data_processed %>% group_by(sex, diet_cond) %>% summarise(mean = mean(total_weight_loss), sd = sd(total_weight_loss)) )
```

```{r Summary Statistics for BMI_2, message = FALSE, echo = FALSE}
#Code for the plots generated in the report for question 2. They're not necessary for the analysis but since I've used these plots for reporting I've included the code here. Strictly for visualization purposes, not included in the knitted file for section 1. 

#Calculated values for male regular diet consumers 
male_reg_diet_values <- diet_data_processed %>% filter(sex=='M',diet_cond=='regular') %>% summarise(mean = mean(starting_bmi), sd = sd(starting_bmi), mean_end = mean(ending_bmi), sd_end = sd(ending_bmi), mean_wt_loss = mean(total_weight_loss), sd_wt_loss = sd(total_weight_loss))
#Calculated values for male no chocolate diet consumers
male_no_choc_diet_values <- diet_data_processed %>% filter(sex=='M',diet_cond=='no_chocolate') %>% summarise(mean = mean(starting_bmi), sd = sd(starting_bmi), mean_end = mean(ending_bmi), sd_end = sd(ending_bmi),mean_wt_loss = mean(total_weight_loss), sd_wt_loss = sd(total_weight_loss))
#Calculated values for female regular diet consumers
female_reg_diet_values <- diet_data_processed %>% filter(sex=='F',diet_cond=='regular') %>% summarise(mean = mean(starting_bmi), sd = sd(starting_bmi), mean_end = mean(ending_bmi), sd_end = sd(ending_bmi),mean_wt_loss = mean(total_weight_loss), sd_wt_loss = sd(total_weight_loss))
#Calculated values for female no chocolate diet consumers
female_no_choc_diet_values <- diet_data_processed %>% filter(sex=='F',diet_cond=='no_chocolate') %>% summarise(mean = mean(starting_bmi), sd = sd(starting_bmi), mean_end = mean(ending_bmi), sd_end = sd(ending_bmi),mean_wt_loss = mean(total_weight_loss), sd_wt_loss = sd(total_weight_loss))

#Starting BMI
#Plotting the starting BMI for all consumers and further plotting the mean and SD for the same
ques2_plot1 <- ggplot(diet_data_processed) + geom_histogram(aes(x=starting_bmi, color = sex), binwidth=1) +
  facet_grid(sex~diet_cond, labeller = as_labeller(Labels))  + 
  labs(x="Starting BMI", y="Frequency", title = "Summary Statistics for Starting BMI") + 
  
#Plotting for males on the regular diet plan
  geom_vline(data=filter(diet_data_processed, sex=="M", diet_cond=='regular'), aes(xintercept=male_reg_diet_values$mean), colour="orange", size = 1)+ 
  geom_text(mapping = aes(x = male_reg_diet_values$mean+3,y = 20, label = paste("Mean =",as.character(round(male_reg_diet_values$mean,1)))), data=filter(diet_data_processed, sex=="M", diet_cond=='regular'), size = 2.5) +
  
  geom_point(data=filter(diet_data_processed, sex=="M", diet_cond=='regular'), aes(x=15, y=20), colour="black") + 
  geom_text(data=filter(diet_data_processed, sex=="M", diet_cond=='regular'), aes(label=paste("SD =", as.character(round(male_reg_diet_values$sd,1))), x=15, y=19), size=2.5)+
  
#Plotting for males on no chocolate diet plan
  geom_vline(data=filter(diet_data_processed, sex=="M", diet_cond=='no_chocolate'), aes(xintercept=male_no_choc_diet_values$mean), colour="orange", size = 1)+ 
  geom_text(mapping = aes(x = male_no_choc_diet_values$mean+3,y = 20, label = paste("Mean =",as.character(round(male_no_choc_diet_values$mean,1)))), data=filter(diet_data_processed, sex=="M", diet_cond=='no_chocolate'), size = 2.5) +
  
  geom_point(data=filter(diet_data_processed, sex=="M", diet_cond=='no_chocolate'), aes(x=15, y=20), colour="black") + 
  geom_text(data=filter(diet_data_processed, sex=="M", diet_cond=='no_chocolate'), aes(label=paste("SD =", as.character(round(male_no_choc_diet_values$sd,1))), x=15, y=19), size=2.5) +

#Plotting for females on the regular diet plan
  geom_vline(data=filter(diet_data_processed, sex=="F", diet_cond=='regular'), aes(xintercept=female_reg_diet_values$mean), colour="orange", size = 1)+ 
  geom_text(mapping = aes(x = female_reg_diet_values$mean+3,y = 20, label = paste("Mean =",as.character(round(female_reg_diet_values$mean,1)))), data=filter(diet_data_processed, sex=="F", diet_cond=='regular'), size = 2.5) +
  
  geom_point(data=filter(diet_data_processed, sex=="F", diet_cond=='regular'), aes(x=15, y=20), colour="black") + 
  geom_text(data=filter(diet_data_processed, sex=="F", diet_cond=='regular'), aes(label=paste("SD =", as.character(round(female_reg_diet_values$sd,1))), x=15, y=19), size=2.5)+
  
#Plotting for females on no chocolate diet plan
  geom_vline(data=filter(diet_data_processed, sex=="F", diet_cond=='no_chocolate'), aes(xintercept=female_no_choc_diet_values$mean), colour="orange", size = 1)+ 
  geom_text(mapping = aes(x = female_no_choc_diet_values$mean+3,y = 20, label = paste("Mean =",as.character(round(female_no_choc_diet_values$mean,1)))), data=filter(diet_data_processed, sex=="F", diet_cond=='no_chocolate'), size = 2.5) +
  
  geom_point(data=filter(diet_data_processed, sex=="F", diet_cond=='no_chocolate'), aes(x=15, y=20), colour="black") + 
  geom_text(data=filter(diet_data_processed, sex=="F", diet_cond=='no_chocolate'), aes(label=paste("SD =", as.character(round(female_no_choc_diet_values$sd,1))), x=15, y=19), size=2.5) 

#Ending BMI
#Plotting the ending BMI for all consumers and further plotting the mean and SD for the same
ques2_plot2 <- ggplot(diet_data_processed) + geom_histogram(aes(x=ending_bmi, color = sex), binwidth=1) +
  facet_grid(sex~diet_cond, labeller = as_labeller(Labels)) + 
  labs(x="Ending BMI", y="Frequency", title = "Summary Statistics for Ending BMI") + 
  
#Plotting for males on the regular diet plan
  geom_vline(data=filter(diet_data_processed, sex=="M", diet_cond=='regular'), aes(xintercept=male_reg_diet_values$mean_end), colour="orange", size = 1)+ 
  geom_text(mapping = aes(x = male_reg_diet_values$mean_end+3,y = 20, label = paste("Mean =",as.character(round(male_reg_diet_values$mean_end,1)))), data=filter(diet_data_processed, sex=="M", diet_cond=='regular'), size = 2.5) +
  
  geom_point(data=filter(diet_data_processed, sex=="M", diet_cond=='regular'), aes(x=15, y=20), colour="black") + 
  geom_text(data=filter(diet_data_processed, sex=="M", diet_cond=='regular'), aes(label=paste("SD =", as.character(round(male_reg_diet_values$sd_end,1))), x=15, y=19), size=2.5)+
  
#Plotting for males on no chocolate diet plan
  geom_vline(data=filter(diet_data_processed, sex=="M", diet_cond=='no_chocolate'), aes(xintercept=male_no_choc_diet_values$mean_end), colour="orange", size = 1)+ 
  geom_text(mapping = aes(x = male_no_choc_diet_values$mean_end+3,y = 20, label = paste("Mean =",as.character(round(male_no_choc_diet_values$mean_end,1)))), data=filter(diet_data_processed, sex=="M", diet_cond=='no_chocolate'), size = 2.5) +
  
  geom_point(data=filter(diet_data_processed, sex=="M", diet_cond=='no_chocolate'), aes(x=15, y=20), colour="black") + 
  geom_text(data=filter(diet_data_processed, sex=="M", diet_cond=='no_chocolate'), aes(label=paste("SD =", as.character(round(male_no_choc_diet_values$sd_end,1))), x=15, y=19), size=2.5) +

#Plotting for females on the regular diet plan
  geom_vline(data=filter(diet_data_processed, sex=="F", diet_cond=='regular'), aes(xintercept=female_reg_diet_values$mean_end), colour="orange", size = 1)+ 
  geom_text(mapping = aes(x = female_reg_diet_values$mean_end+3,y = 20, label = paste("Mean =",as.character(round(female_reg_diet_values$mean_end,1)))), data=filter(diet_data_processed, sex=="F", diet_cond=='regular'), size = 2.5) +
  
  geom_point(data=filter(diet_data_processed, sex=="F", diet_cond=='regular'), aes(x=15, y=20), colour="black") + 
  geom_text(data=filter(diet_data_processed, sex=="F", diet_cond=='regular'), aes(label=paste("SD =", as.character(round(female_reg_diet_values$sd_end,1))), x=15, y=19), size=2.5)+
  
#Plotting for females on no chocolate diet plan
  geom_vline(data=filter(diet_data_processed, sex=="F", diet_cond=='no_chocolate'), aes(xintercept=female_no_choc_diet_values$mean_end), colour="orange", size = 1)+ 
  geom_text(mapping = aes(x = female_no_choc_diet_values$mean_end+3,y = 20, label = paste("Mean =",as.character(round(female_no_choc_diet_values$mean_end,1)))), data=filter(diet_data_processed, sex=="F", diet_cond=='no_chocolate'), size = 2.5) +
  
  geom_point(data=filter(diet_data_processed, sex=="F", diet_cond=='no_chocolate'), aes(x=15, y=20), colour="black") + 
  geom_text(data=filter(diet_data_processed, sex=="F", diet_cond=='no_chocolate'), aes(label=paste("SD =", as.character(round(female_no_choc_diet_values$sd_end,1))), x=15, y=19), size=2.5) 

#WeightLoss
#Plotting the weight loss for all consumers and further plotting the mean and SD for the same
ques2_plot3 <- ggplot(diet_data_processed) + geom_histogram(aes(x=total_weight_loss, color = sex), binwidth=1) +
  facet_grid(sex~diet_cond, labeller = as_labeller(Labels))  + 
  labs(x="Weight Loss", y="Frequency", titles = "Summary Statistics for Weight Loss") + 
  
#Plotting for males on the regular diet plan
  geom_vline(data=filter(diet_data_processed, sex=="M", diet_cond=='regular'), aes(xintercept=male_reg_diet_values$mean_wt_loss), colour="orange", size = 1)+ 
  geom_text(mapping = aes(x = male_reg_diet_values$mean_wt_loss+3,y = 20, label = paste("Mean =",as.character(round(male_reg_diet_values$mean_wt_loss,1)))), data=filter(diet_data_processed, sex=="M", diet_cond=='regular'), size = 2.5) +
  
  geom_point(data=filter(diet_data_processed, sex=="M", diet_cond=='regular'), aes(x=-11, y=20), colour="black") + 
  geom_text(data=filter(diet_data_processed, sex=="M", diet_cond=='regular'), aes(label=paste("SD =", as.character(round(male_reg_diet_values$sd_wt_loss,1))), x=-11, y=19), size=2.5)+
  
#Plotting for males on no chocolate diet plan
  geom_vline(data=filter(diet_data_processed, sex=="M", diet_cond=='no_chocolate'), aes(xintercept=male_no_choc_diet_values$mean_wt_loss), colour="orange", size = 1)+ 
  geom_text(mapping = aes(x = male_no_choc_diet_values$mean_wt_loss+3,y = 20, label = paste("Mean =",as.character(round(male_no_choc_diet_values$mean_wt_loss,1)))), data=filter(diet_data_processed, sex=="M", diet_cond=='no_chocolate'), size = 2.5) +
  
  geom_point(data=filter(diet_data_processed, sex=="M", diet_cond=='no_chocolate'), aes(x=-11, y=20), colour="black") + 
  geom_text(data=filter(diet_data_processed, sex=="M", diet_cond=='no_chocolate'), aes(label=paste("SD =", as.character(round(male_no_choc_diet_values$sd_wt_loss,1))), x=-11, y=19), size=2.5) +

#Plotting for females on the regular diet plan
  geom_vline(data=filter(diet_data_processed, sex=="F", diet_cond=='regular'), aes(xintercept=female_reg_diet_values$mean_wt_loss), colour="orange", size = 1)+ 
  geom_text(mapping = aes(x = female_reg_diet_values$mean_wt_loss+3,y = 20, label = paste("Mean =",as.character(round(female_reg_diet_values$mean_wt_loss,1)))), data=filter(diet_data_processed, sex=="F", diet_cond=='regular'), size = 2.5) +
  
  geom_point(data=filter(diet_data_processed, sex=="F", diet_cond=='regular'), aes(x=-11, y=20), colour="black") + 
  geom_text(data=filter(diet_data_processed, sex=="F", diet_cond=='regular'), aes(label=paste("SD =", as.character(round(female_reg_diet_values$sd_wt_loss,1))), x=-11, y=19), size=2.5)+
  
#Plotting for females on no chocolate diet plan
  geom_vline(data=filter(diet_data_processed, sex=="F", diet_cond=='no_chocolate'), aes(xintercept=female_no_choc_diet_values$mean_wt_loss), colour="orange", size = 1)+ 
  geom_text(mapping = aes(x = female_no_choc_diet_values$mean_wt_loss+3,y = 20, label = paste("Mean =",as.character(round(female_no_choc_diet_values$mean_wt_loss,1)))), data=filter(diet_data_processed, sex=="F", diet_cond=='no_chocolate'), size = 2.5) +
  
  geom_point(data=filter(diet_data_processed, sex=="F", diet_cond=='no_chocolate'), aes(x=-11, y=20), colour="black") + 
  geom_text(data=filter(diet_data_processed, sex=="F", diet_cond=='no_chocolate'), aes(label=paste("SD =", as.character(round(female_no_choc_diet_values$sd_wt_loss,1))), x=-11, y=19), size=2.5) 
```

---

#Question 3
```{r Visualization of Post Diet BMI, message = FALSE}
#R code associated for the third request in the assignment
#Visualizing the overall distribution of post diet BMIs
( ques3_plot1 <- ggplot(diet_data_processed, aes(x=ending_bmi, y=..density..)) + geom_histogram(binwidth=1, alpha = 0.5) +
  geom_density()+
  labs(x="End of Trial BMIs", y="Density", title = "Visualization Of Overall Post Diet BMIs")+
  geom_vline(aes(xintercept=mean(ending_bmi)), colour="orange", size = 1)+ 
  geom_text(aes(x = mean(ending_bmi)+1, y = 0.19, label = "Mean"), size = 2.5, color = "black", show.legend = FALSE))

#Visualizing the distributions of post diet BMIs for each gender and diet category by plotting a histogram and density plot
(ques3_plot2 <- ggplot(diet_data_processed, aes(x=ending_bmi, y=..density.., color = sex)) + geom_histogram(binwidth=1) +
  facet_grid(sex~diet_cond, labeller = as_labeller(Labels)) + 
  geom_density()+
  labs(x="End of Trial BMIs", y="Density", title = "Visualization Of Post Diet BMIs By Gender And Type of Diet")+
  
  #plotting of mean post diet BMIs for each gender and diet category
  geom_vline(data=filter(diet_data_processed, sex=="M", diet_cond=='regular'), aes(xintercept=male_reg_diet_values$mean_end), colour="orange", size = 1)+ 
  geom_text(data=filter(diet_data_processed, sex=="M", diet_cond=='regular'), mapping = aes(x = male_reg_diet_values$mean_end+3, y = 0.19, label = "Mean"), size = 2.5, color = "black", show.legend = FALSE) +
  
geom_vline(data=filter(diet_data_processed, sex=="M", diet_cond=='no_chocolate'), aes(xintercept=male_no_choc_diet_values$mean_end), colour="orange", size = 1)+ 
  geom_text(data=filter(diet_data_processed, sex=="M", diet_cond=='no_chocolate'), mapping = aes(x = male_no_choc_diet_values$mean_end+3, y = 0.19, label = "Mean"), size = 2.5, color = "black", show.legend = FALSE) + 
  
  geom_vline(data=filter(diet_data_processed, sex=="F", diet_cond=='regular'), aes(xintercept=female_reg_diet_values$mean_end), colour="orange", size = 1)+ 
  geom_text(data=filter(diet_data_processed, sex=="F", diet_cond=='regular'), mapping = aes(x = female_reg_diet_values$mean_end+3, y = 0.19, label = "Mean"), size = 2.5, color = "black", show.legend = FALSE) +
  
geom_vline(data=filter(diet_data_processed, sex=="F", diet_cond=='no_chocolate'), aes(xintercept=female_no_choc_diet_values$mean_end), colour="orange", size = 1)+ 
  geom_text(data=filter(diet_data_processed, sex=="F", diet_cond=='no_chocolate'), mapping = aes(x = female_no_choc_diet_values$mean_end+3, y = 0.19, label = "Mean"), size = 2.5, color = "black", show.legend = FALSE))
```

---

#Question 4 
```{r T-test Comparing Diet Plans, message = FALSE}
#Reviewing diet data by weight loss and diet plan
( diet_data_by_weightloss <- diet_data_processed %>% group_by(diet_cond) %>% 
  summarise(mean = mean(total_weight_loss), sd = sd(total_weight_loss, na.rm = TRUE), Count = n()) )

#Since there is a difference in the mean of weight loss for each type of diet plan, plotting a histogram to have a look at how different their distributions are 
( ques4_plot1 <- ggplot(diet_data_processed, aes(total_weight_loss,..density.., fill = diet_cond)) + geom_histogram(binwidth = 1, alpha = 0.5, position = "identity")+ labs(x = "Weight Loss", y = "Density", title = "Weight Loss By Diet Plan") )

#Distributions of weight loss for each diet plan is different, hence performing t-test for further analysis
t.test(total_weight_loss ~ diet_cond, data = diet_data_processed)
```

---

# Section 2

#Report

1) A plot showing the relationship between impulsivity and age.

```{r echo=FALSE, message = FALSE}
print(ques1_plot1)
```

The above scatterplot confirms the suspicion held by the board that the relationship between age and impulsivity for the sample is non-linear. The loess lines created for the sample males and females are curved, thus indicating that a change in age is not directly proportional to a change in impulsivity. For example - The plot clearly displays how with increasing age, increasing impulsivity is not necessary, hence this confirms that the relationship is non-linear.

```{r echo=FALSE, message = FALSE}
print(ques1_plot2)
```

Upon plotting the relationship by gender, it can further be deduced that the relationship for both males and females is non-linear but a pattern is observed. The above plots display how with age there tends to be an increase in impulsivity for women - this is further established by viewing the data points pertaining to females as most lie above the loess line. Where as for men it is observed that with age the impulsivity score is more or less constant with a slight decline in impulsivity score appearing for men above the age of 50, however the score seems to rise once again for men aged 70. 

---

2) Summary statistics for BMI at the end of the trial

```{r echo = FALSE, message = FALSE}
print(ques2_plot3)
```

The above summary statistics for weight loss reveal that while some men and women that chose the regular diet achieved weight loss the average weight loss for the people that took up the regular diet is still very low. Hence the regular diet plan had little success in helping people through their weight loss journey when compared to the no chocolate diet. 

```{r echo = FALSE, message = FALSE}
print(ques2_plot1)
```

```{r echo = FALSE, message = FALSE}
print(ques2_plot2)
```

After calculating the summary statistics for the starting and ending BMIs separately for each sex and diet plan, it can be concluded that men and women that took up the no chocolate diet had a healthier mean BMI therefore they lost more weight than the men and women that took up the regular diet. The average BMIs for the men and women that took up the regular diet have a very marginal change post diet compared to their pre diet BMIs. 

---

3) Visualization of the distribution of Post-Diet BMI

```{r echo = FALSE, message = FALSE}
print(ques3_plot1)
```

The above plot clearly conveys the shape and pattern of the entire distribution. It can be concluded that the end of trial BMIs when plotted follow a normal distribution i.e the post diet BMIs for the sample are symmetric around the average and the major concentration of BMIs lie around the average rather than two or three standard deviations away from it. Further the density plot shows that most people had a post diet BMI falling in the range of 20-25. 

Comparison of post diet BMIs for each gender and diet plan is given below - 

```{r echo = FALSE, message = FALSE}
print(ques3_plot2)
```

---

4) t-test comparing diet plans

From the sample the average weight loss for people that took up the no chocolate diet is 3.17 whereas the average weight loss for people that took up the regular diet is -0.38. The mean weight loss is significantly larger for people that took up the no chocolate diet, Welch t(275) = 7.07, p<0.00001. Since the p value is extremely small we can conclude that the average for both diet plans are significantly different and since their difference in averages is not zero we can reject the null hypothesis. 

In all likelihood the weight loss associated with each diet plan will belong to a separate normal distribution indicating that there is a difference between both populations (people who took up the no chocolate diet and people who took up the regular diet). This is displayed in the graph below - 

```{r echo = FALSE, message = FALSE}
print(ques4_plot1)
```

---